Metric Spaces with Expensive Distances

### 2019-04-25 paper/draft_esa.tex
Page breaks
Changes in abstract and intro

We consider two standard operations on metric spaces,
namely the construction of a $1+\eps$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

>>>>

We consider two standard operations on metric spaces,
namely the construction of a $(1+\eps)$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

### 2019-04-25 paper/expensive-distances-main.tex
Page breaks
Rescale images
Changes in abstract and intro

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances that have been computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
runs faster than naive computation of all pairwise distances or runs slightly slower, depending on the approximation
quality.

@@

>>>>

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated the \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
We demonstrate experimentally that on a collection
of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\newpage
